---
title: "Statistical analysis"
author: "Rubén Sánchez Fernández"
lang: en # language,  en: english (default), es: español, ca: catalan, ...
date: '`r format(Sys.Date(),"%d, %B, %Y")`'
output:
  pdf_document:
    toc: yes
    toc_depth: 2
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)

```

```{r libraries, include=FALSE}
# Install packages
# Load packages
# ...

library(knitr)
```


```{r input, include=FALSE}
# Input / Output variables
# Tuning parameters
# ...

```

>This project is created as an exercise for _Analytical software for statistical analysis_ course of Bioinformatics and Biostatistics MSc. (UOC)    


#Statistical analysis

In this report, we will perform a statistical analysis to a biomedical dataset.



##The dataset

The data belongs to the _Duke University Cardiovascular Disease Databank_ and includes records and features of patients under diagnostic cardiac catheterization. The dataset has 3504 patients and 6 features. 

Data obtained from [http://biostat.mc.vanderbilt.edu/DataSets](http://biostat.mc.vanderbilt.edu/DataSets)



##Data overview

We will start by importing the data.

```{r}
library(Hmisc)
getHdata(acath)
```

Let's check we imported the righ dataset.

```{r}
dim(acath)
```

The dataset has 3504 samples and 6 features, as we introduced in the previous section.


Let's use the _summary_, _str_ and _names_ functions to overview the data.

```{r}
names(acath)
```

```{r}
summary(acath)
```

```{r}
str(acath)
```

From the overview, we can notice that we have several _missing values_, specially in _choleste_ and _tvdlm_ features. Let's study the features more closely.   


For instance, let's plot the distribution of _cad.dur_ variable, which is the duration of the heart symptoms.

```{r}
hist(acath$cad.dur, col="darkseagreen1", main="Distribution of heart symptoms \n in the coronary artery disease", xlab="Time", ylab="Frequency")
```

We could also check the maximum and minimum of the variable _age_.

```{r}
print(paste0("The older patient is ", max(acath$age), " years old"))

print(paste0("and the youngest patient is ", min(acath$age), " years old"))
```

Another interesting question we could make is how many patients had high level of cholesterol (more than 200).

```{r}
nrow(subset(acath, acath$choleste>200))
```

1602 patients had high cholesterol.   


##Probability and simulation

Now, let's use probaility theory to answer some questions:

* What is the probability to find more than 60 patients with the disease if 100 tests were performed?

This case follows a binomial distribution, therefore we need to find the probability to have the disease from the data.

```{r}
est<-subset(acath,acath$sigdz==1) #patients with the disease
p<-nrow(est)/nrow(acath) #probability = patients with the disease/total patients
```

Once we have the probability _p_, we can solve the question:

```{r}
p60<-1-pbinom(60,100,p,lower.tail=T) #dist binomial
print(paste0("The probability to find more than 60 patients with the disease after evaluating 100 tests is ", p60))
```


* Probability of not having the disease if someone has a narrow coronary artery?

This follows conditional probability as:
$$P(no\:disease|narrow\:artery)=\frac{P(no\:disease\bigcap narrow\:artery)}{P(narrow\:artery)}$$

Let's calculate the probabilities:

```{r}
sya<-subset(acath,acath$tvdlm==0 & acath$sigdz==1)
psya<-nrow(sya)/nrow(acath) #probability no disease and narrow artery
ps<-psya/p #conditional probability
print(paste0("Probability of not having the disease while having narrow coronary artery is ",ps))
```


In some situations, performing simulations can be very useful when analyzing data. For instance, let's simulate 10,000 times the number of patients with a narrow coronary artery if 100 tests were performed.

```{r}
set.seed(12345)
t<-rbinom(10000,100,p)
str(t)
#plots
par(mfrow=c(1,2))
hist(t,col="lightblue")
plot(density(t))
mean(t)
sd(t)
```


As we could expect, the mean obtained simulating 10,000 times is very close to the 1-time mean. It is interesting that the standard deviation is not important, therefore there is not much variability in the process.


##Regression analysis

Regression analysis is also very useful when answering questions about our data. For instance, let's question how age influences in the duration of the symptoms.

```{r}
pairs(acath)
```

Let's generate the linear regression model with _symptoms duration_ and _age_.

```{r}
RegModel <- lm(cad.dur~age, data=acath)
summary(RegModel)
```

Determination coefficient is 0.08226. The model has a low explanatory power, so the model is not idoneal or the variables are independent.

Next, let's calculate the correlation coefficient between these variables.

```{r}
cor(acath[,c("age","cad.dur")])
```

The coefficient is 0.2868. As we suspected, the dependency is low.

After this analysis, we can conclude that variable _age_ doesn't influence in a significant level to the variable _symptoms duration_.

